---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# duckspatial

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/duckspatial)](https://CRAN.R-project.org/package=duckspatial)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/Cidree/duckspatial/graph/badge.svg)](https://app.codecov.io/gh/Cidree/duckspatial)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

**duckspatial** is an R package that simplifies the process of reading and writing vector spatial data (e.g., `sf` objects) in a [DuckDB](https://duckdb.org/) database. This package is designed for users working with geospatial data who want to leverage DuckDB’s fast analytical capabilities while maintaining compatibility with R’s spatial data ecosystem.

## Installation

You can install the development version of duckspatial from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("Cidree/duckspatial")
```

## Example

This is a basic example which shows how to set up DuckDB for spatial data manipulation, and how to write/read vector data.

```{r}
library(duckdb)
library(duckspatial)
library(sf)
```


First, we create a connection with a DuckDB database (in this case in memory database), and we make sure that the spatial extension is installed, and we load it:

```{r connect}
## create connection
conn <- dbConnect(duckdb())

## install and load spatial extension
ddbs_install(conn)
ddbs_load(conn)
```

Now we can get some data to insert into the database. We are creating 10,000,000 random points.

```{r}
## create n points
n <- 10000000
random_points <- data.frame(
  id = 1:n,
  x = runif(n, min = -180, max = 180),  # Random longitude values
  y = runif(n, min = -90, max = 90)     # Random latitude values
)

## convert to sf
sf_points <- st_as_sf(random_points, coords = c("x", "y"), crs = 4326)

## view first rows
head(sf_points)
```

Now we can insert the data into the database using the `ddbs_write_vector()` function. We use the `proc.time()` function to calculate how long does it take, and we can compare it with writing a shapefile with the `write_sf()` function:

```{r}
## write data monitoring processing time
start_time <- proc.time()
ddbs_write_vector(conn, sf_points, "test_points")
end_time <- proc.time()

## print elapsed time
elapsed_duckdb <- end_time["elapsed"] - start_time["elapsed"]
print(elapsed_duckdb)
```

```{r}
## write data monitoring processing time
start_time <- proc.time()
shpfile <- tempfile(fileext = ".shp")
write_sf(sf_points, shpfile)
end_time <- proc.time()

## print elapsed time
elapsed_shp <- end_time["elapsed"] - start_time["elapsed"]
print(elapsed_shp)
```

In this case, we can see that DuckDB was `r round(elapsed_shp / elapsed_duckdb, 1)` times faster. Now we will do the same exercise but reading the data back into R:

```{r}
## write data monitoring processing time
start_time <- proc.time()
sf_points_ddbs <- ddbs_read_vector(conn, "test_points", crs = 4326)
end_time <- proc.time()

## print elapsed time
elapsed_duckdb <- end_time["elapsed"] - start_time["elapsed"]
print(elapsed_duckdb)
```

```{r}
## write data monitoring processing time
start_time     <- proc.time()
sf_points_ddbs <- read_sf(shpfile)
end_time       <- proc.time()

## print elapsed time
elapsed_shp <- end_time["elapsed"] - start_time["elapsed"]
print(elapsed_shp)
```


For reading, we get a factor of `r round(elapsed_shp / elapsed_duckdb, 1)` times faster for DuckDB. Finally, don't forget to disconnect from the database:

```{r}
dbDisconnect(conn)
```

